---
title: "COVID-19 in Sweden"
description: |
  Analysis of the epidemic stage in Sweden.
author:
  - name: Pian Li
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, echo=FALSE,eval=TRUE,message=FALSE,include=FALSE,warning=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(here)
library(lubridate)
library(DT)
library(kableExtra)

```


```{r out.width = "80%", fig.align = "center"}
library(grid)
library(png)
img <- readPNG(here("images/Flag_of_Sweden.png"))
grid.raster(img)
```


# Data description

There are two datasets used to analyze the change of disease-related data(cases, deaths, mortality rate) and the efficiency of vaccinations during the COVID-19 phase in Sweden. The data of cases and deaths is obtained from [the World Health Organization (WHO)](https://covid19.who.int/table), while data on COVID-19 vaccinations is sourced from the publicly available [Github repository](https://github.com/owid/covid-19-data/tree/master/public/data/vaccinations) created by [Our World in Data (OWID)](https://ourworldindata.org/).


## Structure of data

All variables included in the two datasets are listed below:

```{r echo=FALSE,eval=TRUE}
data_structure <- data.frame(Variables = c("Country",
                         "Date_reported",
                         "New_cases",
                         "New_deaths",
                         "total_vaccinations",
                         "people_vaccinated",
                         "people_fully_vaccinated"),
           Types = c("character",
                     "date",
                     "double",
                     "double",
                     "double",
                     "double",
                     "double"),
           Description = c("name of the country",
                           "date of the observation",
                           "new confirmed cases of COVID-19",
                           "new deaths attributed to COVID-19",
                           "total number of doses administered,each individual dose is counted",
                           "total number of people who received at least one vaccine dose.",
                           "total number of people who received all doses prescribed by the vaccination protocol"))


kable(data_structure, caption = "Data Structure of chosen datasets") %>%
  kable_paper(bootstrap_options = "striped",full_width = F)

```


# Data Analysis

## Cases and deaths

```{r read-data}
vac <- read_csv(here("./data/vaccinations.csv"))
covid <- read_csv(here("./data/WHO-COVID-19-global-data.csv"))
```


```{r tidycovid}
Date_reported = col_date()

covid_tidy_cases <- covid %>% 
  mutate(date = dmy(Date_reported)) %>% 
  mutate_at(vars(date), funs(year, month, day)) %>%
  mutate(Month = month.abb[as.numeric(month)]) %>%
  group_by(year,Month) %>%
  select(year, Month, Country,New_cases,New_deaths) %>%
  filter(Country == 'Sweden') %>%
  tally(New_cases) %>%
  rename(New_cases = n,
         Year = year)

covid_tidy_deaths <- covid %>% 
  mutate(date = dmy(Date_reported)) %>% 
  mutate_at(vars(date), funs(year, month, day)) %>%
  mutate(Month = month.abb[as.numeric(month)]) %>%
  group_by(year,Month)%>%
  select(year, month, Country,New_cases,New_deaths) %>%
  filter(Country == 'Sweden') %>%
  tally(New_deaths) %>%
  rename(New_deaths = n,
         Year = year)


cases_deaths <- left_join(covid_tidy_cases,covid_tidy_deaths) %>%
  mutate("Mortality rate (%)" = round(New_deaths/New_cases * 100,2)) %>%
  rename("Cases" = New_cases,
         "Deaths" = New_deaths)

```

```{r deathrate}
DT::datatable(cases_deaths, caption = "Number of cases and deaths of COVID19 in Sweden", options = list(pageLength = 10,
                                           initComplete = JS(
    "function(settings, json) {",
    "$(this.api().table().header()).css({'background-color': 'black', 'color': '#fff'});",
    "}")
))

```

**Change in cases and deaths**

The table above shows the monthly change of cases and deaths in Sweden from 2020 to 2021. The number of cases and deaths has surged since March of 2020, the largest number of cases and deaths for the entire year recorded in November, around 183000 cases and 2500 deaths reported. 

According to the National Strategy, the test capacity in Sweden has increased substantially during 2021 (April, May and June), in order to lower the unrecorded cases and deaths (Public health Agency of Sweden, 2020). It can be seen that the number of cases and deaths peaked in April 2021, case number then dropped gradually in the following months.


## Vaccinations

```{r tidyvac}

vac_total <- vac %>%
  filter(location == 'Sweden') %>%
  rename(Country = location) %>%
  mutate(date = ymd(date)) %>% 
  mutate_at(vars(date), funs(year, month, day)) %>%
  mutate(Month = month.abb[as.numeric(month)]) %>%
  group_by(Month,year) %>%
  select(Month, Country,total_vaccinations,people_vaccinated,people_fully_vaccinated,daily_vaccinations) %>%
  tally(total_vaccinations) %>%
  rename(Total_vaccinations = n,
         Year = year) %>%
  select(Year,Month,Total_vaccinations) 


vac_fully <- vac %>% 
  filter(location == 'Sweden') %>%
  rename(Country = location) %>%
  mutate(date = ymd(date)) %>% 
  mutate_at(vars(date), funs(year, month, day)) %>%
  mutate(Month = month.abb[as.numeric(month)]) %>%
  group_by(Month,year) %>%
  select(Month, Country,total_vaccinations,people_fully_vaccinated,people_vaccinated,daily_vaccinations) %>%
  tally(people_fully_vaccinated) %>%
  rename(People_fully_vaccination = n,
         Year = year) %>%
  select(Year,Month,People_fully_vaccination) 

vac_one <- vac %>%
  filter(location == 'Sweden') %>%
  rename(Country = location) %>%
  mutate(date = ymd(date)) %>% 
  mutate_at(vars(date), funs(year, month, day)) %>%
  mutate(Month = month.abb[as.numeric(month)]) %>%
  group_by(Month,year) %>%
  select(Month, Country,total_vaccinations,people_fully_vaccinated,people_vaccinated,daily_vaccinations) %>%
  tally(people_vaccinated) %>%
  rename(People_vaccinated = n,
         Year = year) %>%
  select(Year,Month,People_vaccinated) 

vac_tidy <- left_join(vac_total,vac_one) 

vac_tidy_final <- left_join(vac_tidy,vac_fully) %>%
  mutate("Fully vaccinated rate (%)" = round((People_fully_vaccination / Total_vaccinations) * 100, 2)) %>%
  rename("People fully vaccination" = People_fully_vaccination,
         "Total vaccinations" = Total_vaccinations,
         "People vaccinated" = People_vaccinated)

datatable(vac_tidy_final,caption = "Vaccinations in Sweden", options = list(pageLength = 10,
                                           initComplete = JS(
    "function(settings, json) {",
    "$(this.api().table().header()).css({'background-color': 'black', 'color': '#fff'});",
    "}")
))



```


**Mortality rate and Vaccination**

From the table of cases and deaths, the monthly death rate has significantly risen from the beginning of the pandemic, likely resulting from the imperfect treatment system for a new disease. The highest death rate of 15.5% was recorded in April 2020. It is also clear that the mortality rate has fallen from the beginning of 2021, dropped under 1% from February 2021. 

The population in Sweden has received one and all vaccine doses are recorded and shown in the table above. In addition, there was an upward trend in the fully vaccinated rate from the beginning of 2021, almost 44% have been fully vaccinated in August 2021.

Therefore, it is evident that vaccinations effectively reduce the death rate due to the decline in mortality rate and rise in vaccinations during the same period. Sweden should continue the vaccination roll-out, and people should actively participate.


## Summary Statistics

The statistical summary table below is created using the package _arsenal_, which consists of three variables: Cases, Deaths and Mortality rate (%).

```{r statistic,results="asis"}
library(arsenal)

cases_deaths_stat <- cases_deaths %>%
  select(Year, `Cases`,`Deaths`,`Mortality rate (%)`)

table_one <- tableby(Year ~ .,data = cases_deaths_stat,numeric.stats = c("meansd","range"))
summary(table_one,title = "Summary statistic for cases and deaths in Sweden")

```

In Sweden, the testing has increased since April 2021, more and more people can be tested, and the government can obtain accurate data for epidemic prevention and control. Thus, We should notice that the mean cases is much higher in 2021. Additionally, the average deaths and mortality rate in 2021 are below that in 2020, with around 630 deaths and a 0.6% death rate, which means the vaccine can effectively control the spread of the pandemic and lower mortality.


# Reference

Yihui Xie, Joe Cheng and Xianying Tan (2021). DT: A Wrapper of the JavaScript Library ‘DataTables’. R package version 0.18. https://CRAN.R-project.org/package=DT

Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686

Kirill Müller (2020). here: A Simpler Way to Find Your Files. R package version 1.0.1. https://CRAN.R-project.org/package=here

Hao Zhu (2021). kableExtra: Construct Complex Table with 'kable' and Pipe Syntax. R package version 1.3.4. https://CRAN.R-project.org/package=kableExtra

Garrett Grolemund, Hadley Wickham (2011). Dates and Times Made Easy with lubridate. Journal of Statistical Software, 40(3), 1-25. URL https://www.jstatsoft.org/v40/i03/.

Ethan Heinzen, Jason Sinnwell, Elizabeth Atkinson, Tina Gunderson andGregory Dougherty (2021). arsenal: An Arsenal of 'R' Functions for Large-Scale Statistical Summaries. R package version 3.6.3. https://CRAN.R-project.org/package=arsenal

Hannah Ritchie, Edouard Mathieu, Lucas Rodés-Guirao, Cameron Appel, Charlie Giattino, Esteban Ortiz-Ospina, Joe Hasell, Bobbie Macdonald, Diana Beltekian and Max Roser (2020) - "Coronavirus Pandemic (COVID-19)". Published online at OurWorldInData.org. Retrieved from: 'https://ourworldindata.org/coronavirus' [Online Resource]

World Health Organisation (2021). WHO Coronavirus (COVID-19) Dashboard. Retrieved from https://covid19.who.int/

Mathieu, E., Ritchie, H., Ortiz-Ospina, E. et al. A global database of COVID-19 vaccinations. Nat Hum Behav (2021). https://doi.org/10.1038/s41562-021-01122-8

The Public health Agency of Sweden, (2020), Statistics- How many people are ill from COVID-19 in Sweden?, accessed 1.Sep, 2021, https://www.folkhalsomyndigheten.se/the-public-health-agency-of-sweden/communicable-disease-control/covid-19/covid-19-faq/statistics/?exp=72722#_72722






